Univalent Universes and Completness of Reversible Programs: \\ Featherweight HoTT

### 2017-02-11 upi.tex
more details on upi and accompanying code

@@

\newcommand{\alt}{~\mid~}
\newcommand{\patht}[1]{\textsc{PATHS}(#1,#1)}
\newcommand{\fpatht}[1]{\textsc{FREEPATHS}(#1,\Box)}
\newcommand{\fpathp}[2]{\textsc{freepath}~#1~#2}
\newcommand{\pathind}[2]{\textsc{pathind}~#1~#2}
\newcommand{\invc}[1]{!\;#1}
\newcommand{\evalone}[2]{eval(#1,#2)}
\newcommand{\evalbone}[2]{evalB(#1,#2)}
\newcommand{\reflp}{\textsc{refl}}
\newcommand{\reflh}{\mathit{refl}_{\sim}}
\newcommand{\symh}[1]{\mathit{sym}_{\sim}~#1}
\newcommand{\transh}[2]{\mathit{trans}_{\sim}~#1~#2}
\newcommand{\reflq}{\mathit{refl}_{\simeq}}
\newcommand{\symq}[1]{\mathit{sym}_{\simeq}~#1}
\newcommand{\transq}[2]{\mathit{trans}_{\simeq}~#1~#2}
\newcommand{\isequiv}[1]{\mathit{isequiv}(#1)}
\newcommand{\idc}{\mathit{id}_{\boolt}}
\newcommand{\swapc}{\mathit{swap}_{\boolt}}
\newcommand{\assocc}{\mathit{assoc}}
\newcommand{\invl}{\mathit{invl}}
\newcommand{\invr}{\mathit{invr}}
\newcommand{\invinv}{\mathit{inv}^2}
\newcommand{\idlc}{\mathit{idl}}
\newcommand{\idrc}{\mathit{idr}}
\newcommand{\swapswap}{\swapc^2}
\newcommand{\compsim}{\compc_{\isotwo}}
\newcommand{\iso}{\leftrightarrow}
\newcommand{\isotwo}{\Leftrightarrow}
\newcommand{\piso}{\multimapdotbothB~~}
\newcommand{\zt}{\mathbb{0}}
\newcommand{\ot}{\mathbb{1}}
\newcommand{\fc}{\mathit{false}}
\newcommand{\tc}{\mathit{true}}
\newcommand{\boolt}{\mathbb{B}}
\newcommand{\univ}{\mathcal{U}}
\newcommand{\uzero}{\mathcal{U}_0}
\newcommand{\uone}{\mathcal{U}_1}
\newcommand{\Rule}[2]{
\makebox{
$\displaystyle
\frac{\begin{array}{l}#1\\\end{array}}
{\begin{array}{l}#2\\\end{array}}$}}
\newcommand{\proves}{\vdash}
\newcommand{\jdgg}[3]{#1 \proves #2 : #3}
\newcommand{\jdg}[2]{\proves #1 : #2}
\newcommand{\jdge}[3]{\proves #1 = #2 : #3}
%% codes
%% denotations

@@

\title{Univalent Universes and Completness of Reversible Programs}

@@

@@

\[\begin{array}{rcl}
\tau &::=& \mathbb{2} \\
\\
v &::=& \begin{array}[t]{lrcl}
                    & \fc &:& \mathbb{2} \\
              \alt & \tc &:& \mathbb{2}
               \end{array} \\
\\
c &::=& \begin{array}[t]{lrcl}
              & \id &:& \tau \iso \tau \\
               \alt & \swap &:& \mathbb{2} \iso \mathbb{2} \\
               \alt & \circ &:& (\tau_1 \iso \tau_2) \rightarrow (\tau_2 \iso \tau_3)
                              \rightarrow (\tau_1 \iso \tau_3)
               \end{array} \\
\\
! &:& (\tau_1 \iso \tau_2) \rightarrow (\tau_2 \iso \tau_1) \\
\invc{\id} &=& \id \\
\invc{\swap} &=& \swap \\
\invc{(\compc{c_1}{c_2})} &=& \compc{\invc{c_2}\;}{\;\invc{c_1}} \\
\\
\alpha &::=& \begin{array}[t]{lrcl}
               & \id &:& c \isotwo c \\
               \alt & \assocc &:& \compc{c_1}{(\compc{c_2}{c_3})} \isotwo
                                              \compc{(\compc{c_1}{c_2})}{c_3} \\
               \alt & \idlc &:& \compc{\id}{c} \isotwo c \\
               \alt & \idrc &:& \compc{c}{\id} \isotwo c \\
               \alt & \invl &:& \compc{c\;}{\;\invc{c}} \isotwo \id \\
               \alt & \invr &:& \compc{\invc{c}}{c} \isotwo \id \\
               \alt & \bullet &:& (c_1 \isotwo c_2) \rightarrow (c_2 \isotwo c_3)
                                            \rightarrow (c_1 \isotwo c_3) \\
               \alt & \odot &:& (c_1 \isotwo c_1') \rightarrow (c_2 \isotwo c_2')
                                            \rightarrow (\compc{c_1}{c_2} \isotwo \compc{c_1'}{c_2'}) \\
             \end{array}
\end{array}\]
The 2-combinators $\alpha$ also have inverses \ldots

@@

Now we claim that $\mathcal{U}_\Pi$ is equivalent to $\mathcal{U}$ where $\mathcal{U}_\Pi$ is a universe which includes types $\tau$ as a points, combinators $c : \tau_1 \iso \tau_2$ as 1-paths, and 2-combinators $\alpha : c_1 \isotwo c_2$ as 2-paths.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%
\section{Generalizations}

>>>>

%% Amr
%% words to remember :-)
%% sublime unfathomable
%% ---

@@

\newcommand{\Typ}{\mathbf{Type}}
\newcommand{\alt}{~\mid~}
\newcommand{\patht}[1]{\textsc{PATHS}(#1,#1)}
\newcommand{\fpatht}[1]{\textsc{FREEPATHS}(#1,\Box)}
\newcommand{\fpathp}[2]{\textsc{freepath}~#1~#2}
\newcommand{\pathind}[2]{\textsc{pathind}~#1~#2}
\newcommand{\invc}[1]{!\;#1}
\newcommand{\evalone}[2]{eval(#1,#2)}
\newcommand{\evalbone}[2]{evalB(#1,#2)}
\newcommand{\reflp}{\textsc{refl}}
\newcommand{\reflh}{\mathit{refl}_{\sim}}
\newcommand{\symh}[1]{\mathit{sym}_{\sim}~#1}
\newcommand{\transh}[2]{\mathit{trans}_{\sim}~#1~#2}
\newcommand{\reflq}{\mathit{refl}_{\simeq}}
\newcommand{\symq}[1]{\mathit{sym}_{\simeq}~#1}
\newcommand{\transq}[2]{\mathit{trans}_{\simeq}~#1~#2}
\newcommand{\isequiv}[1]{\mathit{isequiv}(#1)}
\newcommand{\idc}{\mathit{id}_{\boolt}}
\newcommand{\swapc}{\mathit{swap}_{\boolt}}
\newcommand{\assocc}{\mathit{assoc}}
\newcommand{\invl}{\mathit{invl}}
\newcommand{\invr}{\mathit{invr}}
\newcommand{\invinv}{\mathit{inv}^2}
\newcommand{\idlc}{\mathit{idl}}
\newcommand{\idrc}{\mathit{idr}}
\newcommand{\swapswap}{\swapc^2}
\newcommand{\compsim}{\compc_{\isotwo}}
\newcommand{\iso}{\leftrightarrow}
\newcommand{\isotwo}{\Leftrightarrow}
\newcommand{\piso}{\multimapdotbothB~~}
\newcommand{\zt}{\mathbb{0}}
\newcommand{\ot}{\mathbb{1}}
\newcommand{\fc}{\mathit{false}}
\newcommand{\tc}{\mathit{true}}
\newcommand{\boolt}{\mathbb{B}}
\newcommand{\univ}{\mathcal{U}}
\newcommand{\uzero}{\mathcal{U}_0}
\newcommand{\uone}{\mathcal{U}_1}
\newcommand{\Rule}[2]{
\makebox{
$\displaystyle
\frac{\begin{array}{l}#1\\\end{array}}
{\begin{array}{l}#2\\\end{array}}$}}
\newcommand{\proves}{\vdash}
\newcommand{\jdgg}[3]{#1 \proves #2 : #3}
\newcommand{\jdg}[2]{\proves #1 : #2}
\newcommand{\jdge}[3]{\proves #1 = #2 : #3}
%% codes
%% denotations

@@

\title{Univalent Universes and Completness of Reversible Programs: \\
  Featherweight HoTT}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Univalence}

Let $X : \Typ$. We have:
\[
\textit{transport-equiv} : \Pi_{P : X \to \Typ}~ \Pi_{x,y:X}~
    x \equiv y  \to P x \simeq P y
\]
For a fibration with base $X$, define:
\[\begin{array}{l}
\textit{is-univalent-fibration} : (X \to \Typ) \to \Typ \\
\textit{is-univalent-fibration}~P =
    \Pi_{x,y:X}~\textit{is-equiv} (\textit{transport-equiv}~P~x~y)
\end{array}\]

\noindent\paragraph*{Examples}
\begin{itemize}
\item Take $X = \Typ$ and $P = \mathit{Id}_{\Typ}$. The statement
  $\textit{is-univalent-fibration}~P$ reduces to $\Pi_{x,y:\Typ}~x \equiv y \to
  x \simeq y$ which is the usual definition of univalence.
\item Take $X = \mathbb{1}$ with element $\star$ and
  $P = \lambda \_. \mathbb{1}$. The statement
  $\textit{is-univalent-fibration}~P$ reduces to
  $(\star\equiv\star)\simeq(\mathbb{1}\simeq\mathbb{1})$ which is true.
\item Take $X = \mathbb{1}$ and $P = \lambda \_. \mathbb{0}$. The statement
  $\textit{is-univalent-fibration}~P$ reduces to
  $(\star\equiv\star)\simeq(\mathbb{0}\simeq\mathbb{0})$ which is also true.
\item Take $X = \mathbb{1}$. For no $P$ other than the above two instances will
  we have a univalent fibration.
\item Take $X = \mathbb{2}$ with elements $\fc$ and $\tc$ and
  $P = \lambda \_. \mathbb{1}$. The statement
  $\textit{is-univalent-fibration}~P$ reduces to $(\fc \equiv \tc) \simeq
  (\mathbb{1}\simeq\mathbb{1})$ which is false.
\item Take $X = \mathbb{2}$ with elements $\fc$ and $\tc$ and
  $P = \lambda \{ \fc \to \mathbb{0}; \tc \to \mathbb{1} \}$ The statement
  $\textit{is-univalent-fibration}~P$ reduces to $(\fc \equiv \tc) \simeq
  (\mathbb{0}\simeq\mathbb{1})$ which is true. That is the only choice of $P$
  that works (modulo symmetry).
\item For any $n > 2$, taking $X$ as sum of $n$ copies of $\mathbb{1}$ can never
  give a univalent fibration.
\end{itemize}

\begin{theorem}
  Let $X$ be path-connected. Then $X$ is the base of a univalent fibration only
  if there exists an $F$ such that $X \simeq \Sigma_{Y : \Typ} \| Y \equiv F \|$
\end{theorem}

\noindent In particular $\Sigma_{Y : \Typ} \| Y \equiv \mathbb{2} \|$ is the base of a
univalent fibration. In some sense this is the simplest univalent universe, our
featherweight HoTT that we aim to characterize as a reversible programming
language based on type isomorphisms.

@@

\[\begin{array}{rcl}
\tau &::=& \mathbb{2} \\
\\
v &::=& \begin{array}[t]{lrcl}
                    & \fc &:& \mathbb{2} \\
              \alt & \tc &:& \mathbb{2}
               \end{array} \\
\\
c &::=& \begin{array}[t]{lrcl}
              & \id &:& \tau \iso \tau \\
               \alt & \swap &:& \mathbb{2} \iso \mathbb{2} \\
               \alt & \circ &:& (\tau_1 \iso \tau_2) \to (\tau_2 \iso \tau_3)
                              \to (\tau_1 \iso \tau_3)
               \end{array} \\
\\
! &:& (\tau_1 \iso \tau_2) \to (\tau_2 \iso \tau_1) \\
\invc{\id} &=& \id \\
\invc{\swap} &=& \swap \\
\invc{(\compc{c_1}{c_2})} &=& \compc{\invc{c_2}\;}{\;\invc{c_1}} \\
\\
\alpha &::=& \begin{array}[t]{lrcl}
               & \id &:& c \isotwo c \\
               \alt & \assocc &:& \compc{c_1}{(\compc{c_2}{c_3})} \isotwo
                                              \compc{(\compc{c_1}{c_2})}{c_3} \\
               \alt & \idlc &:& \compc{\id}{c} \isotwo c \\
               \alt & \idrc &:& \compc{c}{\id} \isotwo c \\
               \alt & \invl &:& \compc{c\;}{\;\invc{c}} \isotwo \id \\
               \alt & \invr &:& \compc{\invc{c}}{c} \isotwo \id \\
               \alt & \bullet &:& (c_1 \isotwo c_2) \to (c_2 \isotwo c_3)
                                            \to (c_1 \isotwo c_3) \\
               \alt & \odot &:& (c_1 \isotwo c_1') \to (c_2 \isotwo c_2')
                                            \to (\compc{c_1}{c_2} \isotwo \compc{c_1'}{c_2'}) \\
             \end{array}
\end{array}\]

The 2-combinators $\alpha$ also have inverses. The above forms a groupoid (see
accompanying code).

@@

Now we claim that $\mathcal{U}_\Pi$ is equivalent to $\mathcal{U}$ where
$\mathcal{U}_\Pi$ is a universe which includes types $\tau$ as a points,
combinators $c : \tau_1 \iso \tau_2$ as 1-paths, and 2-combinators $\alpha : c_1
\isotwo c_2$ as 2-paths. The proof is in a meta-logic which is itself univalent.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%
\section{Generalization I}

Use all finite types instead of just Bool

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%
\section{Generalization II}

Add a HIT to the univalent universe, perhaps something to do with fractionals.
